---
title: "Analysis of Central Arctic Ocean (CAO) prokaryotes using Scanning electron microscopy (SEM) and -energy dispersive X-ray spectroscopy (SEM-EDX)"
author:
  name: "Ashish Verma"
  email: "asve0014@gmail.com"
date: "2026-01-10"
format:
  html:
    toc: true
    code-fold: true
    theme: cosmo

execute:
  echo: true
  warning: false
  message: false
  dir: project   

editor: visual
---

# Load packages and libraries

```{r}
packages <- c("readxl","tidyverse","ggplot2","broom","viridis",
    "patchwork","cowplot","rstatix","zoo","scales","ggpubr","magick","grid","FSA",
    "dunn.test")

#Install missing packages

for (pkg in packages) {
if (!require(pkg, character.only = TRUE)) {
install.packages(pkg)
}
}

# Load packages without printing output
purrr::walk(packages, library, character.only = TRUE, quietly = TRUE)
```

# Percentage distribution of cell shapes

@fig-4a shows the percentage distribution of cell shape between the epipelagic and mesopelagic depth layers.

```{r}
##Read a specific sheet by name
sem <- read_excel("./data/rawdata.xlsx", sheet = "cellcounts")

##Count cells
cells_counts <- sem |>
  group_by(shape, depth_category) |>
  summarise(total = sum(n), .groups = "drop")
```

```{r}
##Recoding the depth layers setting order
cells_counts_rec <- sem |>  #rec = recode
  mutate(
    depth_category =recode(
      depth_category,
      "Upper" = "Epipelagic\n(n=177)",
      "Lower" = "Mesopelagic\n(n=52)"
    ),
    depth_category = factor(
      depth_category,
      levels = c("Epipelagic\n(n=177)", "Mesopelagic\n(n=52)")
    )
  )
```

```{r}
##Percentage of cells
cells_shape_perc <- cells_counts_rec |>
  group_by(depth_category, shape) |>
  summarise(total = sum(n), .groups = "drop") |>
  group_by(depth_category) |>
  mutate(shape_percentage = round((total / sum(total)) * 100, 1))
```

```{r}
#Order the cell shape based on percentage
cells_shape_perc$shape <- factor(
  cells_shape_perc$shape,
  levels = c("Rods", "Cocci", "Vibrios", "Spirals", "Spirochetes")
)
```

```{r}
##Gradient percentage
cells_shape_perc <- cells_shape_perc |>
  mutate(label_color = ifelse(shape_percentage > 23, "white", "black"))  # Black and white colors were set to see the % numbers on contasts
```

```{r}
#| label: fig-4a
#| fig-cap: "Cell shape percentage distribution between epipelgiac and mesopelagic"

##Gradient plot of cell shapes
Fig4a <- ggplot(cells_shape_perc,
                aes(x = depth_category, y = shape, fill = shape_percentage)) +
  geom_tile() +
  geom_text(aes(label = paste0(round(shape_percentage, 1), "%"),
                color = label_color), size = 2) +
  scale_color_identity() +
  scale_fill_viridis_c(option = "cividis",
                       direction = -1,
                       name = "Percentage (%)") +
  labs(fill = "Percentage (%)", x = "Depth", y = "Cell shape") +
  theme_minimal() +
  theme(
    text = element_text(family = "sans"),
    axis.title.x = element_text(colour = "Black", size = 7),
    axis.title.y = element_text(colour = "Black", size = 7),
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 7, face = "bold"),
    panel.grid = element_blank(),
    panel.border = element_rect(
      color = "black",
      linewidth = 1.5,
      fill = NA
    ),
    axis.text = element_text(size = 7, face = "bold"),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1,
      vjust = 1,
      margin = margin(t = 0)
    ),
    plot.title = element_text(hjust = 0.03),
    strip.text = element_text(size = 7, face = "bold"),
    panel.spacing = unit(0.2, "lines")
  )
Fig4a

##save in figures folder
ggsave(
  filename = "Fig4a.png",
  plot = Fig4a,
  path = "figures",
  width = 80,
  height = 80,
  units = "mm",
  dpi = 400
)
```

# SEM image showing morphological features

@fig-4b shows the different morphological features of Arctic prokaryotes seen in scanning electron microscopy

```{r}
#| label: fig-4b
#| fig-cap: "Scanning electron micrographs shows the different morphological features of CAO prokaryotes"

##upload the SEM image
SEM <- image_read("./figures/Fig4b.png")

##Convert to grob for ggplot
g_img <- rasterGrob(SEM, interpolate = TRUE)

##Wrap in a ggplot panel
Fig4b <- ggplot() +
  annotation_custom(g_img) +
  theme_minimal()

##Show the figure
Fig4b
```

# Statistical tests: specific to cell shape between depths

Cell counts for each shape were grouped into epipelagic (upper) and mesopelagic (lower) layers, and the percentage of each cell shape was calculated within each depth category.

```{r}
##To test if there is significant difference between overall cell shape percentage between depths
##Check normality of data in cell shape percentage

##Is percentage distribution different between Epipelagic and Mesopelagic overall, pooling all shapes together?

##Read a specific sheet by name
shape_depth <- read_excel("./data/rawdata.xlsx", sheet = "oceanbasin_depth_percentage")

##Shapiro Wilk test based on depth
##1.Null hypothesis(H0): data follows normal distribution
##2.Alternate hypothesis(H1): data does not follow normal distribution
normality_depth <- shape_depth |>
  group_by(depth) |>
  summarise(n = n(),
            shapiro_p = if (n() >= 3)
              shapiro.test(percentage)$p.value
            else
              NA) |>
  ungroup()
```

```{r}
##p-value < 0.05, Reject Null hypothesis: Data is not normally distributed
#Apply non parametric test

# Overall Mann–Whitney test
wilcox_depth <- wilcox.test(percentage ~ depth,
                        data = shape_depth,
                        exact = FALSE,
                        correct = FALSE)
print(wilcox_depth)
```

```{r}
##There is significant difference in overall cell shape proportions
##check if there is any siginficant difference in a specific shape between two depths

##Check Shapiro–Wilk tests for normality
normality_shape_depth <- shape_depth |>
  group_by(shape, depth) |>
  summarise(
    n      = n(),
    shapiro_p = if (n >= 3) {
      shapiro.test(percentage)$p.value
    } else {
      NA_real_
    },
    .groups = "drop"
  )
print(normality_shape_depth)
```

```{r}
##Based on normality test, data is non-normal (p<0.05), so test with non parametric test

##Wilcoxon per shape: Epipelagic vs Mesopelagic
wilcox_shape_depth <- shape_depth |>
  group_by(shape) |>
  summarise(
    n_ep = sum(depth == "Epipelagic", na.rm = TRUE),
    n_ms = sum(depth == "Mesopelagic", na.rm = TRUE),

    p_value = wilcox.test(
      percentage ~ depth,
      data = pick(everything()),
      exact   = FALSE,
      correct = FALSE
    )$p.value,

    .groups = "drop"
  ) |>
  mutate(p_adj = p.adjust(p_value, method = "bonferroni"))

print(wilcox_shape_depth)
##p-adjusted values shows that there is no siginficant difference between rods, cocci, spirals, vibrios, and spirochete between depths
```

# Statistical tests: specific to cell shape between ocean basins

```{r}
##For ocean_basin-Normality test
normality_test_oceanbasin <- shape_depth |>
  group_by(ocean_basin, depth) |>
  summarise(n = n(),
            shapiro_p = if (n() >= 3)
              shapiro.test(percentage)$p.value
            else
              NA) |>
  ungroup()
print(normality_test_oceanbasin)
```

```{r}
##In most of the cases, p < 0.05, so data is non-normal
##So, KW test done-for comparing two or more independent samples-3 ocean basins here
##OverallKruskall wallis test (ignoring Shape)
kw_oceanbasin <- kruskal.test(percentage ~ ocean_basin, data = shape_depth)
print(kw_oceanbasin)

##No significant difference found
```

# Chi-square test: Finding association of morphological features between epipelagic and mesopelagic layers

```{r}
##Chi square test on morphological features

##Read a specific sheet by name
morph_features <- read_excel("./data/rawdata.xlsx", sheet = "features")
```

```{r}
##Convert columns to factors
morph_features$ocean_basin <- as.factor(morph_features$ocean_basin)
morph_features$depth <- as.factor(morph_features$depth)
morph_features$features <- as.factor(morph_features$features)
morph_features$criteria <- as.factor(morph_features$criteria)
```

```{r}
##Chi-square test function for depth and features
perform_chisq_test <- 
  function(df, feature_name) {
  ##Filter data for the specific feature
  feature_df <- morph_features |>
    filter(features == feature_name)
  ##Create a 2x2 contingency table using depth and criteria
  tbl <-
    xtabs(counts ~ depth + criteria, data = feature_df)
  cat("Contingency Table for Feature:", feature_name, ":\n")
  print(tbl)
  
  ##Check if the table is 2x2
  if (nrow(tbl) == 2 && ncol(tbl) == 2) {
    ##Perform chi-square test without correction
    chisq_result <- suppressWarnings(chisq.test(tbl, correct = FALSE))
    cat("\nExpected Frequencies for Feature:", feature_name, ":\n")
    print(chisq_result$expected)
    
    ##Decision logic for test selection
    if (any(chisq_result$expected <= 5)) {
      ##Fisher's Exact Test: Used when any expected frequency is <= 5
      ##Provides exact p-values and is preferred for small expected frequencies
      cat("\nExpected frequency <= 5, using Fisher's Exact Test...\n")
      fisher_result <- fisher.test(tbl)
      print(fisher_result)
    } else {
      ##Chi-Square Test without correction: Applied when all expected frequencies > 5
      ##Reliable and straightforward when expected counts are sufficient
      cat("\nExpected frequency > 5, using standard Chi-Square Test...\n")
      chisq_result <- suppressWarnings(chisq.test(tbl, correct = FALSE))
      print(chisq_result)
    }
  } else {
    cat("\nInvalid 2x2 table for feature:", feature_name, "\n")
  }
}
```

# Chi-square test results

All ten morphological features observed in Arctic prokaryotes were evaluated using chi-square tests to examine their association with epipelagic and mesopelagic depth layers. For this analysis, data from both depth layers were combined and compared.

```{r}
##Unique features list
distinct_features <- unique(morph_features$features)

##Print overall results first
cat("\n=================== Overall Chi-Square Tests ===================\n")
for (features in distinct_features) {
  cat("\n=== Feature:", features, "===\n")
  perform_chisq_test(morph_features, features)
}
```

# Chi-square test results: oceanbasin-specific analysis

This analysis evaluates whether morphological features are significantly associated with individual oceanbasins (Amundsen, Makarov and Nansen) using oceanbasin-specific chi-square tests. The results will appear as ten chi-square analysis under each basin.

```{r}
##Print oceanbasin-specific results
cat("\n=================== ocean_basin-specific Chi-Square tests ===================\n")
for (sb in unique(morph_features$ocean_basin)) {
  seabsin_data <- morph_features |>
    filter(ocean_basin == sb)
  cat("\nocean_basin:", sb, "\n")
  for (feat in distinct_features) {
    cat("\nFeature:", feat, "\n")
    perform_chisq_test(ocean_basin_data, feat)
  }
}
```

# Chi-square table

@fig-4c shows results from the chi-square test analysis.

```{r}
#| label: fig-4c
#| fig-cap: "Chi-square test results showing the association of morphological features with different depth layers of Central Arctic Ocean"

##Create Chi square-table
significant_features <- data.frame(
  ocean_basin = c(
    "Amundsen +\nMakarov +\nNansen (n = 229)",
    "",
    "",
    "Amundsen (n = 65)",
    "Makarov (n = 104)",
    "",
    "",
    "Nansen (n = 60)"
  ),
  `Morphological features` = c("TL + EPS", "VL", "TL", "VL", "TL + EPS", "PL", "TL", "EPS"),
  `Chi- square test (p-value)` = c(
    "9.4 (0.002)",
    "9.5 (0.001)",
    "8.6 (0.003)",
    "< 0.001",
    "12.7 (< 0.001)",
    "16.2 (< 0.001)",
    "11.5 (< 0.001)",
    "4.5 (0.03)"
  ),
  `Odds ratio` = c(
    "2.7 (E/M)",
    "2.8 (M/E)",
    "5.3 (E/M)",
    "28.5 (M/E)",
    "5.1 (E/M)",
    "10.1 (E/M)",
    "7.4 (E/M)",
    "5.0 (E/M)"
  )
)

##Set column names
colnames(significant_features) <- c("Ocean basin", "Features", "Chi-square (p-value)", "Odds ratio")

##Create custom theme with sans font and size 7
nm_style <- ttheme(
  base_size = 7,
  colnames.style = colnames_style(
    face = "bold",
    size = 7,
    color = "black",
    fontfamily = "sans"
  ),
  tbody.style = tbody_style(
    face = "plain",
    size = 6.5,
    color = "black",
    fontfamily = "sans"
  )
)

##Make the table
Fig4c_table <- ggtexttable(significant_features, rows = NULL, theme = nm_style)
Fig4c_table
```

# Elemental analysis of Arctic prokaryotes using SEM-EDX

@fig-4d shows the elemental ratio of Arctic prokaryotes

```{r}
##Fig.4d: Read a specific sheet by name
edx_ratio_cell <- read_excel("./data/rawdata.xlsx", sheet = "edx_cell_mol_ratio")

##Modify Depth labels
edx_ratio_cell$depthID <- factor(edx_ratio_cell$depthID,
                                  levels = c("Epipelagic", "Mesopelagic"))
```

```{r}

##Use pivot_longer to reshape the data
edx_ratio_cell <- edx_ratio_cell |>
  pivot_longer(
    cols = c("C_N", "C_P", "N_P"),
    names_to = "mol_ratio",
    values_to = "value"
  )
```

```{r}
#Calcualte summary statistics
summary_stats_edx <- edx_ratio_cell |>
  group_by(depthID, mol_ratio) |>
  summarise(
    median = median(value, na.rm = TRUE),
    SE = sd (value, na.rm = TRUE) / sqrt(n()),
    n = n(),
    CI = 1.96 * SE
  ) |>
  ungroup()
```

```{r}
# Normality Test (Shapiro-Wilk Test)
shapiro_results <- edx_ratio_cell |>
  group_by(depthID, mol_ratio) |>
  summarise(
    shapiro_p = shapiro.test(value)$p.value,
    .groups = "drop"
  )
print(shapiro_results)
```

```{r}
# C:N ratio (C_N)
test_C_N <- wilcox.test(
  value ~ depthID,
  data = edx_ratio_cell %>% filter(mol_ratio == "C_N")
)
print(test_C_N)

# C:P ratio (C_P)
test_C_P <- t.test(
  value ~ depthID,
  data = edx_ratio_cell %>% filter(mol_ratio == "C_P")
)
print(test_C_P)

# N:P ratio (N_P)
test_N_P <- wilcox.test(
  value ~ depthID,
  data = edx_ratio_cell %>% filter(mol_ratio == "N_P")
)
print(test_N_P)

```

```{r}
##Compute label positions
max_y_per_ratio <- edx_ratio_cell |>
  group_by(mol_ratio) |>
  summarise(max_y = max(value, na.rm = TRUE)) |>
  mutate(
    line_y = max_y * 1.05,
    ##line position
    text_y = max_y * 1.08,
    ##text slightly above line
    label = c("*", "*", "n.s.")  ##adjusted based on statistics calculated above
  )
```

```{r}
#| label: fig-4d
#| fig-cap: "Stiochiometric ratio of different elements in the different depth layers of Central Arctic Ocean"

# Plot violin plot for the ratios, faceted by depthID
violin_plot <- ggplot(edx_ratio_cell,
                aes(x = mol_ratio, y = value, fill = depthID)) +
  geom_violin(
    lwd = 0.6,
    alpha = 0.5,
    trim = TRUE,
    position = position_dodge(width = 0.9)
  ) +
  geom_boxplot(
    width = 0.2,
    position = position_dodge(width = 0.9),
    outlier.shape = NA,
    alpha = 0.7
  ) +
  geom_point(
    aes(fill = depthID),
    position = position_jitterdodge(),
    shape = 21,
    size = 2,
    alpha = 0.2
  ) +
  scale_fill_viridis_d(option = "cividis", direction = -1) +
  facet_wrap( ~ mol_ratio, scales = "free") +
  scale_x_discrete(labels = c(
    "C_N" = "C:N\n(n = 48, 31)",
    "C_P" = "C:P\n(n = 52, 29)",
    "N_P" = "N:P\n(n = 51, 28)"
  )) +
  labs(x = "Stoichiometric ratios", y = "Molar ratio", fill = "Depth") +
  theme_bw() +
  theme(
    text = element_text(family = "sans"),
    legend.position = "top",
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 7),
    panel.grid = element_blank(),
    panel.border = element_rect(
      color = "black",
      linewidth = 1.5,
      fill = NA
    ),
    axis.title.x = element_text(colour = "Black", size = 7),
    axis.title.y = element_text(colour = "Black", size = 7),
    axis.text.x = element_text(
      angle = 0,
      hjust = 0.5,
      size = 7
    ),
    strip.text = element_blank()
  )

###Add significaNce lines 

Fig4d <- violin_plot +
  geom_segment(
    data = max_y_per_ratio,
    aes(
      x = 0.7,
      xend = 1.4,
      y = line_y,
      yend = line_y
    ),
    inherit.aes = FALSE
  ) +
  geom_text(
    data = max_y_per_ratio,
    aes(x = 1.0, y = text_y, label = label),
    size = 3,
    inherit.aes = FALSE
  )

Fig4d

##Save the plot
ggsave(
  "Fig4d.png",
  path = "figures",
  plot = violin_plot,
  width = 110,
  height = 100,
  units = "mm",
  dpi = 400
)
```

# Wrap and combine figures in one plot

@fig-4 shows the combined plots of cell shape, morphological features and elemental composition in CAO prokaryotes

```{r}
#| label: fig-4
#| fig-cap: "Morphology and elemental composition of CAO prokaryotes analysed by SEM and SEM-EDX"
##Wrap it using patchwork
Fig4a_wrap <- wrap_elements(Fig4a)
Fig4b_wrap <- wrap_elements(Fig4b)
##Assemble top and bottom rows
top_row <- Fig4a_wrap + Fig4b_wrap + plot_layout(widths = c(2, 3))
bottom_row <- Fig4c_table + Fig4d

##Combine rows with annotation
Fig4 <- top_row / bottom_row +
  plot_annotation(tag_levels = 'a',
                  theme = theme(plot.tag = element_text(
                    size = 7,
                    face = "bold",
                    family = "sans"
                  )))
Fig4

##save in figures folder as png
ggsave(
  filename= "Fig4_SEM.png",
  plot = Fig4,
  path ="figures",
  width = 180,
  height = 185,
  units = "mm",
  dpi = 500
)

##save in figures folder as pdf
ggsave(
  filename= "Fig4_SEM.pdf",
  plot = Fig4,
  path ="figures",
  width = 180,
  height = 185,
  units = "mm",
  dpi = 500
)

```

# Distribution of morphological features in epipelagic and mesopelagic depth layers

@fig-EFa shows the distribution of morphological features in CAO prokaryotes uisng heatmap gradient

```{r}
##Extended Figure: Plotting yes and no category foR morphological features
##Read a specific sheet by name
morph_feat <- read_excel("./data/rawdata.xlsx", sheet = "features")

##Compute proportions for counts
morph_feat_percentage <- morph_feat |>
  group_by(ocean_basin, depth, feature_acronym) |>
  mutate(total = sum(counts), percentage = counts / total * 100) |>
  ungroup()
```

```{r}
##Ensure Percentage is numeric
morph_feat_percentage$percentage <- as.numeric(as.character(morph_feat_percentage$percentage))

##Modify Depth labels
morph_feat_percentage$depth <- as.character(morph_feat_percentage$depth)
morph_feat_percentage$depth[morph_feat_percentage$depth == "Upper"] <- "Epipelagic"
morph_feat_percentage$depth[morph_feat_percentage$depth == "Lower"] <- "Mesopelagic"
morph_feat_percentage$depth <- factor(morph_feat_percentage$depth, levels = c("Epipelagic", "Mesopelagic"))
head(morph_feat_percentage, 6)
```

```{r}
##Calculate summary statistics: mean only
##Ensure all combinations of MorphologicalFeatures, depth, and criteria are present
morph_feat_summary <- morph_feat_percentage |>
  group_by(depth, feature_acronym, criteria) |>
  summarize(mean_percentage = mean(percentage, na.rm = TRUE),
            .groups = "drop")
```

```{r}
##Heatmap to show plot
##Capture original order
original_order <- unique(morph_feat_summary$feature_acronym)

##List of significant features
significant_features <- c("TL + EPS", "VL", "TL")

##Modify f1_summary Feature labels
morph_feat_summary <- morph_feat_summary |>
  mutate(
    feature_starred = ifelse(
      feature_acronym %in% significant_features,
      paste0("** ", feature_acronym),
      # or "* " / "** " based on threshold
      feature_acronym
    )
  )
```

```{r}
##Match the new feature_starred factor to the original order
morph_feat_summary <- morph_feat_summary |>
  mutate(feature_starred = factor(
    feature_starred,
    levels = ifelse(
      original_order %in% significant_features,
      paste0("** ", original_order),
      original_order
    )
  ))
```

```{r}
#| label: fig-EFa
#| fig-cap: "Distribution of morphological features"

##Plot the heatmap
EFa_SEM <- ggplot(morph_feat_summary,
                  aes(x = criteria, y = feature_starred, fill = mean_percentage)) +
  geom_tile(color = NA) +
  scale_fill_viridis_c(option = "cividis",
                       direction = -1,
                       name = "percentage (%)") +
  facet_grid(. ~ depth) +
  labs(x = "Criteria", y = "Morphological features") +
  theme_bw() +
  theme(
    text = element_text(family = "sans"),
    axis.title.x = element_text(color = "black", size = 8),
    axis.title.y = element_text(color = "black", size = 8),
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 7, face = "bold"),
    axis.text = element_text(size = 7, face = "bold"),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    strip.text = element_text(size = 7, face = "bold"),
    panel.grid = element_blank(),
    panel.border = element_rect(
      color = "black",
      linewidth = 1.5,
      fill = NA
    ),
    panel.spacing = unit(0.2, "lines")
  )
EFa_SEM

##Save the plot as png
ggsave(
  filename = "EFa_SEM.png",
  path = "figures",
  width = 95,
  height = 80,
  units = "mm",
  dpi = 500
)
```

# Molar percentages between prokaryotic cells and vesicle like (VL) structures

@fig-EFb shows the elemental differences between CAO prokaryotic cells and vesicle like structures

```{r}
##Fig.EFb:Plot Prokaryotic cell vs Vesicle like structures in mol %
##Read specific sheet by name
edx_cell_VL <- read_excel("./data/rawdata.xlsx", sheet = "cell_VL")
```

```{r}
##Summary statistcis of mol ratio
summary_edx_cell_VL <- edx_cell_VL |>
  group_by(category, element) |>
  summarise(
    mean = mean(mol_percent, na.rm = TRUE),
    se = sd(mol_percent, na.rm = TRUE) / sqrt (n()),
    n = n(),
    CI = 1.96 * se,
    .groups = "drop"
  )
```

```{r}
##Define the outlier removal function based on IQR
edx_cell_VL_outliers <- function(data) {
  Q1 <- quantile(data, 0.25, na.rm = TRUE)  # First quartile (25th percentile)
  Q3 <- quantile(data, 0.75, na.rm = TRUE)  # Third quartile (75th percentile)
  IQR_value <- Q3 - Q1  # Interquartile range
  
  ##Calculate the bounds for outliers
  lower_bound <- Q1 - 1.5 * IQR_value
  upper_bound <- Q3 + 1.5 * IQR_value
  
  ##Return only values within the bounds, and mark outliers as NA
  cleaned_data <- ifelse(data >= lower_bound &
                           data <= upper_bound, data, NA)
  
  return(cleaned_data)
}
```

```{r}
##Apply outlier removal by category and molpercent and filter out NA values in a single step
edx_cell_VL_no_outliers <- edx_cell_VL |>
  group_by(category, element) |>  # group by category
  mutate(value_no_outliers_mol_percent = edx_cell_VL_outliers(mol_percent)) |>  
  # Apply outlier removal
  filter(!is.na(value_no_outliers_mol_percent)) |>  
  # Remove rows where Value_no_outliers is NA
  ungroup()  # Remove grouping after applying the function
```

```{r}
##Calcualte summary statistics
summary_edx_cell_VL_no_outliers <- edx_cell_VL_no_outliers |>
  group_by(category, element) |>
  summarise(
    median = median(value_no_outliers_mol_percent, na.rm = TRUE),
    SE = sd (value_no_outliers_mol_percent, na.rm = TRUE) / sqrt(n()),
    n = n(),
    CI = 1.96 * SE
  ) |>
  ungroup()
```

```{r}
##Normality Test (Shapiro-Wilk Test)
cell_VL_shapiro_test_C <- shapiro.test(edx_cell_VL_no_outliers$value_no_outliers_mol_percent[edx_cell_VL_no_outliers$element == "C"])
cell_VL_shapiro_test_N <- shapiro.test(edx_cell_VL_no_outliers$value_no_outliers_mol_percent[edx_cell_VL_no_outliers$element == "N"])
cell_VL_shapiro_test_P <- shapiro.test(edx_cell_VL_no_outliers$value_no_outliers_mol_percent[edx_cell_VL_no_outliers$element == "P"])
cell_VL_shapiro_test_Mg <- shapiro.test(edx_cell_VL_no_outliers$value_no_outliers_mol_percent[edx_cell_VL_no_outliers$element == "Mg"])


##View normality test results
cell_VL_shapiro_test_C
cell_VL_shapiro_test_N
cell_VL_shapiro_test_P
cell_VL_shapiro_test_Mg
```

```{r}
##Mann-Whitney U test for each element mol % (since data is not normal)
##For Carbon
cell_VL_mann_whitney_C <- wilcox.test(value_no_outliers_mol_percent ~ category, data = edx_cell_VL_no_outliers[edx_cell_VL_no_outliers$element == "C", ])
print(cell_VL_mann_whitney_C)

##For Nitrogen
cell_VL_mann_whitney_N <- wilcox.test(value_no_outliers_mol_percent ~ category, data = edx_cell_VL_no_outliers[edx_cell_VL_no_outliers$element == "N", ])
print(cell_VL_mann_whitney_N)

##For Phosphorus
cell_VL_mann_whitney_P <- wilcox.test(value_no_outliers_mol_percent ~ category, data = edx_cell_VL_no_outliers[edx_cell_VL_no_outliers$element == "P", ])
print(cell_VL_mann_whitney_P)

##For Magnesium¨

cell_VL_mann_whitney_Mg <- wilcox.test(value_no_outliers_mol_percent ~ category, data = edx_cell_VL_no_outliers[edx_cell_VL_no_outliers$element == "Mg", ])
print(cell_VL_mann_whitney_Mg)
```

```{r}
#| label: fig-EFb
#| fig-cap: "Molar percentages of elements in prokaryotic cell and vesicle like structures"


##Ensure 'element' is a factor with the specified levels
edx_cell_VL_no_outliers$element <- factor(edx_cell_VL_no_outliers$element,
                                                levels = c("C", "N", "P", "Mg", "Ca"))


##Plot boxplot for the mol_percent, faceted by category
EFb_SEM <- ggplot(
  edx_cell_VL_no_outliers,
  aes(x = element, y = value_no_outliers_mol_percent , fill = category)
) +
  geom_violin(
    lwd = 0.6,
    alpha = 0.5,
    trim = TRUE,
    position = position_dodge(width = 0.9)
  ) +
  geom_boxplot(
    width = 0.2,
    position = position_dodge(width = 0.9),
    outlier.shape = NA,
    alpha = 0.7
  ) +
  geom_point(
    aes(fill = category),
    position = position_jitterdodge(),
    shape = 21,
    size = 1,
    alpha = 0.2
  ) +
  scale_fill_viridis_d(option = "cividis", direction = -1) +
  facet_wrap( ~ element, scales = "free") +
  scale_x_discrete(
    labels = c(
      "C" = "C (n = 84, 46)",
      "N" = "N (n = 86, 23)",
      "P" = "P (n = 81, 49)",
      "Ca" = "Ca (n = 32)",
      "Mg" = "Mg (n =11, 43)"
    )
  ) +
  labs(x = "Elements", y = "Mol %", fill = "category") +
  theme_bw() +
  theme(
    text = element_text(family = "sans"),
    legend.position = "top",
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 7),
    panel.grid = element_blank(),
    panel.border = element_rect(
      color = "black",
      linewidth = 1.5,
      fill = NA
    ),
    axis.title.x = element_text(colour = "Black", size = 7),
    axis.title.y = element_text(colour = "Black", size = 7),
    axis.text.x = element_text(
      angle = 0,
      hjust = 0.5,
      size = 7
    ),
    strip.text = element_blank()
  )
EFb_SEM

##Save the plot
ggsave(
  filename = "EFb_edx_cell_VL.png",
  path = "figures",
  plot = EFb_SEM,
  width = 96,
  height = 100,
  units = "mm",
  dpi = 500
)
```

# Combined figures: morphological features and mol% comaprison

@fig-EF shows the distribution of morphological features and mol% comparison

```{r}
#| label: fig-EF
#| fig-cap: "Morphological features and mol% comparison"

##combine figures
ExtendedFig <- EFa_SEM + EFb_SEM + plot_layout(ncol = 2, widths = c(1, 2)) +
  plot_annotation(
    tag_levels = 'a',
    ##gives 'a' and 'b'
    theme = theme(plot.tag = element_text(
      size = 7,
      face = "bold",
      family = "sans" 
    ))
  )
ExtendedFig

##Save the plot as png
ggsave(
  filename = "ExtendedFig.png",
  path = "figures",
  plot = ExtendedFig,
  width = 180,
  height = 120,
  units = "mm",
  dpi = 500
)

##Save the plot as pdf
ggsave(
  filename = "ExtendedFig.pdf",
  path = "figures",
  plot = ExtendedFig,
  width = 180,
  height = 120,
  units = "mm",
  dpi = 500
)
```
